---
title: "joint structural and functional registration with antsr"
author: "Brian B. Avants et al."
date: "5/30/2018"
output: html_document
---

```{r setup, include=FALSE}
library(ANTsR)
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Human connectome project data.

Preprocessing steps involved ( at this instructional level ) for each subject:

1. preprocess fmri
    * distortion correction
    * average bold
    * motion correction
    * rigid map to t1

2. preprocess t1
    * crude brain extraction
    * segmentation
    * map to template
    
3. extract relevant networks
    * default mode
    * motor 
    
4. perform a joint structural / functional registration between these subjects.

##  IO and modalities

```{r io}
rdir = "/Users/stnava/code/structuralFunctionalJointRegistration/"
if ( ! exists("id") ) id = '3026'
if ( ! exists("id") ) id = '2001'
# collect image data
# data/LS2001/unprocessed/3T/T1w_MPR1/LS2001_3T_T1w_MPR1_gdc.nii.gz
t1fn = paste0( rdir, 'data/LS',id, "/unprocessed/3T/T1w_MPR1/LS",id,"_3T_T1w_MPR1_gdc.nii.gz")
# now the bold data
boldfnsL = Sys.glob( paste0( rdir, 'data/LS',id, "/LS", id, "fmri/unprocessed/3T/rfMRI_REST1_*/*REST1_LR_gdc.nii.gz" ) )
boldfnsR = Sys.glob( paste0( rdir, 'data/LS',id, "/LS", id, "fmri/unprocessed/3T/rfMRI_REST1_*/*REST1_RL_gdc.nii.gz" ) )
# get the ref data
reffns = Sys.glob( paste0( rdir, 'data/LS',id, "/LS", id, "fmri/unprocessed/3T/rfMRI_REST1_*/*SBRef_gdc.nii.gz" ) )
```

##  Distortion correction (without a field map)

Let us first "undistort"

```{r}
if ( ! exists( "und" ) ) {
  i1 = antsImageRead( reffns[1] )
  i2 = antsImageRead( reffns[2] )
  und = buildTemplate( i1, list( i1, i2 ), "SyN" )
  } 
t1 = antsImageRead( t1fn ) %>% n3BiasFieldCorrection( 8 ) %>%
  n3BiasFieldCorrection( 4 )
t1rig = antsRegistration( und, t1, "BOLDRigid" )
t1reg = antsRegistration( und, t1, "SyNOnly", initialTransform = t1rig$fwd  )
t1toBOLD = antsApplyTransforms( und, t1, t1reg$fwdtransforms )
```



## Brain masking

Use the BOLD mask to extract the brain from the t1 (for expedience, usually we 
would have run `antsCorticalThickness` already.)

```{r}
bmask = getMask( und )
t1mask = antsApplyTransforms( t1, bmask, t1reg$invtransforms, interpolator = 'nearestNeighbor' )
t1 = n4BiasFieldCorrection( t1, t1mask, 8 ) %>%
  n4BiasFieldCorrection( t1mask, 4 )
plot( t1 * t1mask, axis = 3 )
```


## Tissue segmentation

Now we can segment the T1.

```{r segt1}
# a simple method
################################################
if ( ! exists( "t1seg" ) ) {
  qt1 = iMath( t1, "TruncateIntensity", 0, 0.95 )
  t1seg = kmeansSegmentation( qt1, 3, t1mask, 0.2 ) 
  volumes = labelStats( t1seg$segmentation, t1seg$segmentation )
  rownames( volumes ) = c("background",'csf', 'gm', 'wm' )
  volumes$NormVolume = volumes$Volume / sum( volumes$Volume[-1])
  pander::pander( volumes[ , c("LabelValue","Volume","NormVolume")] )
derka
  # we look and realize this is not good - fix & redo - caused by local hyperintensity
  if ( volumes["wm","NormVolume"] < 0.2 ) {
    t1mask2 = t1mask * thresholdImage( t1seg$segmentation, 1, 2 )
    t1seg = kmeansSegmentation( t1, 3, t1mask2, 0.1 ) 
    }
}
plot( t1, t1seg$segmentation, axis = 3, window.overlay=c(0,3) )
boldseg = antsApplyTransforms( und, t1seg$segmentation, t1reg$fwdtransforms, 
                               interpolator = 'nearestNeighbor' )
plot(und,boldseg,axis=3,alpha=0.5)
########################################
```


## Template mapping

* include prior information e.g. from meta-analysis or anatomy

```{r totemplate}
if ( ! exists( "treg" ) ) {
  data( "powers_areal_mni_itk" )
  myvoxes = 1:nrow( powers_areal_mni_itk )
  anat = powers_areal_mni_itk$Anatomy[myvoxes]
  syst = powers_areal_mni_itk$SystemName[myvoxes]
  Brod = powers_areal_mni_itk$Brodmann[myvoxes]
  xAAL  = powers_areal_mni_itk$AAL[myvoxes]
  if ( ! exists( "ch2" ) )
    ch2 = antsImageRead( getANTsRData( "ch2" ) )
  treg = antsRegistration( t1 * t1mask, ch2, 'SyN' )
}
concatx =  c( treg$invtransforms, t1reg$fwdtransforms )
pts2bold = antsApplyTransformsToPoints( 3, powers_areal_mni_itk, concatx, 
                                        whichtoinvert = c( T, F, T ) )
ptImg = makePointsImage( pts2bold, bmask, radius = 3 )
plot( und, ptImg, axis=3, window.overlay = range( ptImg ) )
###########################################################
```


# Extracting canonical functional network maps


## preprocessing

back to the fmri ...

    * undistort
    * motion correction

then we will be ready for first-level analysis ...

```{r motion}
if ( ! exists( "motcorr" ) ) {
  bold = antsImageRead( boldfnsR )
  avgBold = getAverageOfTimeSeries( bold )
  # map to und 
  boldUndTX = antsRegistration( und, avgBold, "SyN" )
  boldUndTS = antsApplyTransforms( und, bold, boldUndTX$fwd, imagetype = 3  )
  avgBold = getAverageOfTimeSeries( boldUndTS )
  motcorr = antsrMotionCalculation( boldUndTS, verbose = F )
  }
plot( ts( motcorr$fd$MeanDisplacement ) )
print( names( motcorr ) )
````

## network modeling

now we can do some additional level-one analysis to extract relevant networks.

    * default mode
    * motor 

```{r denoise}
# use tissue segmentation to guide compcor 
csfAndWM = thresholdImage( boldseg, 1, 1 ) + ( thresholdImage( boldseg, 3, 3 ) %>% iMath("ME",1))
ccmat = timeseries2matrix( motcorr$moco_img, csfAndWM )
noiseu = compcor( ccmat, ncompcor = 6 )
# convert GM to a matrix
smth = c( 5, 5, 5, 2.0 ) # this is for sigmaInPhysicalCoordinates = T
smth = rep( 1, 4 ) # might need to mess with this some ...
simg = smoothImage( motcorr$moco_img, smth, sigmaInPhysicalCoordinates = F )
gmmat = timeseries2matrix( simg, thresholdImage( boldseg, 2, 2 ) )
tr = antsGetSpacing( bold )[4]
gmmatf = frequencyFilterfMRI( gmmat, tr = tr, freqLo = 0.01, freqHi = 0.1,  opt = "trig" )
# smoothing or filtering?
dfnpts = which( powers_areal_mni_itk$SystemName == 'Default Mode' ) # & 
#  powers_areal_mni_itk$Anatomy == "Posterior Cingulate Gyrus" )
dfnmask = maskImage( ptImg, ptImg, level = dfnpts, binarize = T )
dfnmat = timeseries2matrix( simg, dfnmask )
dfnmatf = frequencyFilterfMRI( dfnmat, tr = tr, freqLo = 0.01, freqHi = 0.1,  opt = "trig" )
dfnsignal = rowMeans( data.matrix( dfnmatf ) )
dfndf = data.frame( signal = dfnsignal, noiseu = noiseu, fd = motcorr$fd )
myform = paste( " mat ~ ", paste( names( dfndf ), collapse = "+") )
dfnmdl = ilr( dfndf, list( mat = data.matrix( data.frame( gmmatf )) ), myform )
####################################
##############
```

now we can look at the full continuous beta map for the default mode network

```{r dfnBetas}
dfnBetaImg = makeImage(  thresholdImage( boldseg, 2, 2 ),  dfnmdl$tValue["signal",] )
plot( und, dfnBetaImg, alpha = 0.98 , axis = 3, window.overlay = c(3, max(dfnBetaImg )),
      nslices = 24, ncolumns = 6 )
ofn = paste0( rdir, "features/LS", id, '_dfnBetaImg.nii.gz' )
antsImageWrite( dfnBetaImg, ofn )
ofn = paste0( rdir, "features/LS", id, '_undistort.nii.gz' )
antsImageWrite( und, ofn )
ofn = paste0( rdir, "features/LS", id, '_t1ToBold.nii.gz' )
antsImageWrite( t1toBOLD, ofn )
```


Now repeat all of this for the next subject by changing the ID.


# Structural functional joint registration

Finally, use all of this data to do a joint registration using both structural 
and functional features.

```{r poorMansHyperalignment}
# antsRegistration with multivariateExtras
f1fn = paste0( rdir, "features/LS", id, '_dfnBetaImg.nii.gz' )
f2fn = paste0( rdir, "features/LS", id, '_undistort.nii.gz' )
f3fn = paste0( rdir, "features/LS", id, '_t1ToBold.nii.gz' )
#############
```

